Effect of experimental habitat connectivity on dung beetle communities

### 2025-08-01 corridor_docs/eric_ms_thesis/EscobarChena_etal_ms.Rmd
reorg of repo for MS

---
title: "The Influence of Connectivity on Dung Beetle Communities"
author: "ERIC ESCOBAR-CHENA"
date: "`r format(Sys.time(), '%d %B, %Y')`"
# date: "\\vspace{-1em}"
# header-includes:
 # - \AtBeginDocument{\let\maketitle\relax} # this removes default title section
#  - \usepackage{lineno}
#  - \linenumbers
#  - \usepackage[default]{sourcesanspro}
#  - \usepackage{fancyhdr} # modify header into left, center, right
#  - \pagestyle{fancy} # cues the use of fancy headers
#  - \fancyfoot{} # cues the use of fancy footer
#  - \fancyhead[L]{EEC MS Thesis Draft} # month & year on header left 
#  - \fancyhead[R]{p. \thepage} # name and page on header right
#  - \usepackage{subcaption}
#  - \usepackage{float}
# # # - \fancyfoot[R]{ }
#  - \usepackage{setspace}

header-includes:
  - \AtBeginDocument{\let\maketitle\relax} # this removes default title section
  - \usepackage{float} #to keep figure in the location placed instead of end
  - \floatplacement{figure}{H}
  - \usepackage{booktabs}
  - \raggedbottom
  - \usepackage{fancyhdr}
  - \pagestyle{plain}
  - \fancyfoot[C]{\thepage}
  - \fancyhead[L]{}
  - \fancyhead[R]{\tiny{EEC MS Thesis Draft (`r format(Sys.time(), '%d %B %Y')`)}}
  - \thispagestyle{fancy}
  - \usepackage{tabu}
  - \usepackage{makecell}
  - \usepackage{pdflscape}
  - \usepackage{longtable}
  - \usepackage{lscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
  - \setlength\parindent{22pt}
  - \usepackage{indentfirst}
  # - \usepackage{caption}
  # - \captionsetup[figure]{name={Figure 1-},labelsep=period}
  # - \captionsetup[table]{name={Table 1-},labelsep=period}
  - \renewenvironment{abstract}{}{} #suppress the abstract
  - \usepackage{setspace}
  - \setcounter{secnumdepth}{0} # removes numbering from sections and subsections
  - \usepackage{sectsty}\allsectionsfont{\raggedright}
  # - \usepackage{sectsty}\sectionfont{\centering}{\normalfont} # center # headings
  # - \usepackage{sectsty}\subsectionfont{\centering}{\normalfont} # center ## headings
  - \usepackage{sectsty}\subsubsectionfont{\bfseries} # center ## headings
  # - \usepackage{sectsty}\subsubsectionfont{\itshape} # no centering but italic subsu
  # - \usepackage{sectsty}\subsubsectionfont{\centering}{ital} # center ### headings
  # - \usepackage{sectsty}\sectionfont{\centering\color{darkmidnightblue}}

 # - doublespacing # manipulate line spacing 
# - \usepackage{parskip} # manipulate indents and spacing before/after paragra phs 
# - \usepackage{xcolor} # changes colors of text
# # http://latexcolor.com/ to choose colors
# # define the colors used above with rgb codes
# - \definecolor{darkcerulean}{rgb}{0.03, 0.27, 0.49}
# - \definecolor{darkmidnightblue}{rgb}{0.0, 0.2, 0.4}
# # to center to top level (#) headings (my name) and change color of font
# # more info # https://www.ctan.org/pkg/sectsty
# - \usepackage{sectsty}
# - \sectionfont{\centering\color{darkmidnightblue}}
# # to center to sub-sub-headings (###) and change color of font
# - \usepackage{sectsty} 
# - \subsectionfont{\centering\color{darkmidnightblue}}
# - \usepackage{sectsty} 
# - \subsubsectionfont{\raggedright\color{darkmidnightblue}}
# - \usepackage{wrapfig} 
# - \newcommand{\blandscape}{\begin{landscape}}
# - \newcommand{\elandscape}{\end{landscape}}
# - \usepackage[font=small,labelfont=bf]{caption} #Required for specifying captions to tables and figures
# - \usepackage{indentfirst}
# - \setlength\parindent{22pt} # uncomment if you want indented paragraphs
bibliography: ["eric-thesis.bib"]
#output:
#  bookdown::pdf_document2:
#    toc: false
#    number_sections: true
#    fig_caption: true
#classoption: 12pt
# output: bookdown::word_document2
# output: word_document
# output: html_document

output: bookdown::pdf_document2
pdf_document:
  includes:
    # in_header: style/preamble.tex
    template: null
    keep_tex: no
    extra_dependencies: ["float"]
    number_sections: true
# documentclass: style/nsf2
biblio-style: apalike
csl: ecology.csl
---

@@

```{r setup, include = FALSE}

@@

```
\newpage

\vspace*{5cm}

\begin{center}
THE INFLUENCE OF CONNECTIVITY ON DUNG BEETLE COMMUNITIES
\end{center}

\vspace{5cm}

\begin{center}
By
\end{center}

\vspace{1cm}

\begin{center}
ERIC ESCOBAR-CHENA
\end{center}

\vspace{5cm}

\begin{center}
A THESIS PRESENTED TO THE GRADUATE SCHOOL\\  
OF THE UNIVERSITY OF FLORIDA IN PARTIAL FULFILLMENT\\ 
OF THE REQUIREMENTS FOR THE DEGREE OF\\  
MASTER OF SCIENCE\\
\end{center}

\begin{center}
UNIVERSITY OF FLORIDA
\end{center}

\begin{center}
2025
\end{center}

\vspace*{11cm}

\begin{center}
© 2025 Eric Escobar-Chena
\end{center}

\vspace*{10cm} 
<!-- _Your dedication is typed here. It should begin with the word “To.” (To my Mom is a typical dedication) If your dedication is longer than a single line it should be single-spaced, and centered vertically and horizontally_ -->

\begin{center}
To my mom, dad, and two sisters. This thesis is a chapter in our immigrant story. 
\end{center}

\begin{center}

\textbf{ACKNOWLEDGMENTS}

\end{center}
<!-- _Acknowledgments must be written in complete sentences. Do not use direct address. For example, instead of Thanks, Mom and Dad!, you should say I thank my parents._  -->
\doublespacing

I thank the USDA Forest Service for maintaining experimental landscapes and assisting in getting established at the site. I also wanted to specifically thank Thomas Smith for his help in data collection, Sara Escobar-Chena for her help in processing and data entry.

\begin{center}  
\textbf{LIST OF ABBREVIATIONS}
\end{center}

\vspace{1cm}

\begin{flushleft}

SRS: Savannah River Site.

GLMM: Generalized Linear Mixed Model.

\end{flushleft}

\vspace*{5cm}

\singlespacing

\begin{center}  
Abstract of Thesis Presented to the Graduate School\\  
of the University of Florida in Partial Fulfillment of the\\  
Requirements for the Degree of Master of Science.\\   
\end{center}

\vspace*{1cm}

\begin{center}
THE INFLUENCE OF CONNECTIVITY ON DUNG BEETLE COMMUNITIES
\end{center}

\begin{center}
By
\end{center}

\begin{center}
Eric Escobar-Chena   
\end{center}

\begin{center}
August 2025   
\end{center}

\begin{flushleft}
Chair: Emilio Bruna\\  
Major: Wildlife Ecology and Conservation\\   
\end{flushleft}

\doublespacing

Habitat fragmentation threatens biodiversity across the globe as habitat loss, isolation, and edge effects become increasingly prevalent. Corridors have become an important tool in order to combat the negative effects of fragmentation, however they are difficult to study in natural systems without incurring confounding effects. To observe changes in insect community composition as an effect of landscape features I sampled dung beetles in a landscape scale experiment. I did not see a difference in species richness or diversity, but dung beetle abundances were higher in continuous forest habitat and open habitat patches connected by a corridors than in isolated patches.

\begin{center}
CHAPTER 1\\  
INTRODUCTION\\  
\end{center}

\doublespacing

Corridors have been shown to be an important mechanism for facilitating the movement of organisms through fragmented landscapes [@haddad_corridor_2003]. These movements are hypothesized to prevent species diversity from declining in fragments, as well as help maintain the ecosystem services provided by these species (at both the patch- and landscape-level)[@burt_ants_2022]. Although there is some evidence that animals disperse between patches via corridors, and that connected patches have higher species diversity than unconnected ones, little work to date has investigated the consequences of these corridor-driven patterns for ecosystem services.

\begin{center}
CHAPTER 2\\  
THE INFLUENCE OF CONNECTIVITY ON DUNG BEETLE COMMUNITIES\\  
\end{center}

\doublespacing

@@

\begin{center} 
\textbf{Methods}
\end{center}
### Study site

@@

\begin{center} 
\textbf{Results}
\end{center}

@@

\begin{center} 
\textbf{Discussion}
\end{center}
<!-- Paragraph 1: Overview: While there may indeed be effects of patch area and connectivity on dung beetle abundance on community composition, other (landscape-scale) drivers appear to be more important.  -->

@@

**[Paragraph 7: spatial scale of experimental plots vs. dung detectability]**

@@

\begin{center}
CHAPTER 3\\  
CONCLUSIONS AND FUTURE DIRECTIONS\\  
\end{center}

\doublespacing 
Moving forward, there is a bounty of knowledge yet to be collected for dung beetles in the south eastern United States. Relative to what is already known in tropical ecosystems, this study provides a glimpse on how dung beetles are responding to fragmentation and connectivity in subtropic pine habitats, but much can still be learned about dung beetles in these spaces and the role they play in ecosystems as well as how they interact with landscapes. The unique role of dung beetles as waste removers is fascinating enough and of great importance for maintaining landscapes so dominated by pastures like Southern US [@cheng_dweller_2022; @WOS:000368243300001]. Not only waste removal but also secondary effect such as seed dispersal and parasitic reduction are important to understand in the context of fragmentation [@fincher_effects_1975; @vulinec_dung_2002]. SRS provides an excellent experimental design for direct comparison of landscapes but another main goal of the site is to study dispersal, dung beetles would be an excellent system for studying movement. They already show promise, in preliminary trials I released beetles from both the connected and rectangular patches with baited traps in the central patch. I recovered one recapture which originated from the connected patch, indicating that individuals do move through corridors.

@@

\begin{center} 
\textbf{References}
\end{center}

\smallskip

@@

<!-- (text - Nathan Fig of plots) -->

<!-- ```{r, label = pitfalls, echo = FALSE, message = FALSE, warning = FALSE,fig.cap = 'Experimental landscapes indicating location of the pitfall traps.',  out.width="25%", fig.align='center'} -->

<!-- a<-readPNG("images/pitfalls.png") -->

<!-- b<-readPNG("images/srs-map.png") -->

<!-- grid.arrange(rasterGrob(a),rasterGrob(b),ncol=1) -->

<!-- ``` -->

```{r echo=FALSE, fig.align='center', fig.cap="All experimental blocks at SRS. Filled squares indicate the blocks in which dung beetles were sampled. Each block has a unique ID number.", message=FALSE, warning=FALSE, label=srs-map, out.width="60%"}

knitr::include_graphics("images/srs-map.png")

```{r, label = patches, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Overhead view of one of the experimental blocks indicating the different types of patches and the distance between them (from Nathan et al. 2018 and reused with permission).", out.width="70%", fig.align='center'}

knitr::include_graphics("images/patches-pic.png")

```{r, label = pitfalls, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Diagram indicating the placement pitfall traps in each patch and in the matrix. Traps in patches were approximately 40 m from the patch edge.", out.width="70%", fig.align='center'}

knitr::include_graphics("images/pitfalls.png")

<!-- ```{r Figure_hill, echo=FALSE,message = FALSE,warning=FALSE} -->

<!-- library(iNEXT) -->

<!-- # -->

<!-- # btl_data<-read_csv(here("corridor_docs","eric_ms_thesis","ms_data","data_clean","clean_btl_counts.csv")) -->

<!-- # spp_codes<-read_csv(here("corridor_docs","eric_ms_thesis","ms_data","data_raw","species_codes.csv")) -->

<!-- btl_sums_patch<-btl_data %>% -->

<!--   pivot_longer(pvin:osyl,names_to = "species",values_to = "n") %>% -->

<!--   group_by(species,patch,block) %>% -->

<!--   mutate(patch=case_when( -->

<!--     patch == "c" ~ "Corridor", -->

<!--     patch == "m" ~ "Matrix", -->

<!--     patch == "w" ~ "Winged", -->

<!--     patch == "r" ~ "Rectangle", -->

<!--     .default = as.character(patch))) %>% -->

<!--   summarize(n=sum(n, na.rm=TRUE)) %>% -->

<!--   arrange(patch,desc(n)) %>% -->

<!--   rename(sp_code=species, -->

<!--          n_patch=n) %>% -->

<!--   mutate(pb=paste(patch,block,sep="-")) %>%  -->

<!--   ungroup() %>%  -->

<!--   select(-patch,-block) %>%  -->

<!--   pivot_wider( -->

<!--     names_from = pb, -->

<!--     values_from = c(n_patch))  -->

<!-- div.data <- btl_sums_patch |> -->

<!--   column_to_rownames(var = "sp_code") -->

<!-- div.btl <- iNEXT(div.data, q = c(0,1,2), datatype = "abundance") -->

<!-- z <- iNEXT(div.data, q=c(0,1,2), datatype="abundance") -->

<!-- ggiNEXT(z, facet.var="Order.q", color.var="Assemblage") + -->

<!--   theme_bw(base_size=18) -->

<!-- ``` -->

\newpage
```{r, label = patch, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Average total dung beetle abundance by patch type with standard deviations as error bars.", out.width="70%", fig.align='center'}

<!-- ```{r sp_richness_patches, echo = FALSE, fig.cap = "Dung beetle secies richness in three different --- and the forest matrix surrounding patches.", out.width = '100%'} -->

<!-- knitr::include_graphics("./corridor_beetles/corridor_docs/eric_ms_thesis/images/sp_richness_patches.png") -->

<!-- ``` -->

@@

@@

<!-- ```{r abundance-patch, echo=FALSE,message = FALSE,warning=FALSE} -->
<!-- # https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html -->

<!-- btl_data<-read_csv(here("corridor_docs","eric_ms_thesis","ms_data","data_clean","clean_btl_counts.csv")) -->
<!-- spp_codes<-read_csv(here("corridor_docs","eric_ms_thesis","ms_data","data_raw","species_codes.csv")) -->

<!-- btl_sums_patch<-btl_data %>% -->
<!--   select(-sum) %>% -->
<!--   pivot_longer(pvin:osyl,names_to = "species",values_to = "n") %>% -->
<!--   group_by(patch) %>% -->
<!--   summarize(n=sum(n, na.rm=TRUE)) %>% -->
<!--   arrange(patch,desc(n)) %>% -->
<!--   mutate(patch=case_when( -->
<!--     patch == "c" ~ "Corridor", -->
<!--     patch == "m" ~ "Matrix", -->
<!--     patch == "w" ~ "Winged", -->
<!--     patch == "r" ~ "Rectangle", -->
<!--     .default = as.character(patch))) -->

<!-- # if wanted to make a bar graph -->
<!-- # ggplot(btl_sums_patch, aes(x=patch, y=n)) +  -->
<!-- #   geom_bar(stat = "identity") -->
<!-- #  -->
<!-- abund_patch <- knitr::kable(btl_sums_patch, -->
<!--       digits = 2, -->
<!--       align="lr", -->
<!--       format="latex", -->
<!--       row.names = FALSE, -->
<!--       booktabs=T, -->
<!--       linesep = "", #removes the blank line after every 5 lines -->
<!--       caption = "Total dung beetles captured in all replicates of a patch type.") %>% -->
<!--   kable_styling(bootstrap_options = c("striped"), -->
<!--                 full_width = F, -->
<!--                 latex_options=c("scale_down","hold_position"), -->
<!--                 font_size = 12, -->
<!--                 position = "center") %>% -->
<!--   # column_spec(1, italic = T) %>% -->
<!--   row_spec(0, bold = T) -->
<!--   # row_spec(1:15, hline_after = T) -->
<!-- abund_patch -->

<!-- ``` -->

<!-- ```{r Table-aovMO, echo=FALSE,message = FALSE,warning=FALSE} -->
<!-- aovM0<-read_csv(here("corridor_docs","eric_ms_thesis","tables","aovM0.csv")) %>% -->
<!--   mutate(term=case_when( -->
<!--     term == "patch_type" ~ "Patch type", -->
<!--     .default = as.character(term))) |> -->
<!-- # aovM0<-read_csv("~/Rstudio/corridor_beetles/corridor_docs/eric_ms_thesis/tables/aovM0.csv") %>%  -->
<!--   kable( -->
<!--       digits = 3, -->
<!--       align="lcc", -->
<!--       format="latex", -->
<!--       row.names = FALSE, -->
<!--       booktabs=T, -->
<!--       linesep = "", #removes the blank line after every 5 lines -->
<!--       caption = "AOV TABLE CAPTION") %>% -->
<!--   kable_styling(bootstrap_options = c("striped"), -->
<!--                 full_width = F, -->
<!--                 latex_options=c("scale_down","hold_position"), -->
<!--                 font_size = 12, -->
<!--                 position = "center") %>% -->
<!--   # column_spec(1, italic = T) %>% -->
<!--   row_spec(0, bold = T) -->
<!--   # row_spec(1:15, hline_after = T) -->
<!-- aovM0 -->
<!-- ``` -->

\newpage
\blandscape
```{r Table-MO, echo=FALSE,message = FALSE,warning=FALSE}
M0<-read_csv(here("corridor_docs","eric_ms_thesis","tables","M0.csv")) %>%
  replace_na(list(group=" ")) %>% 
  rename(SE = std.error,
         Estimate=estimate,
         Term=term,
         Group=group,
         Effect=effect,
         Statistic=statistic,
         P=p.value) %>% 
  mutate(Effect=if_else(Effect=="ran_pars","Random",Effect)) %>% 
  mutate(Effect=if_else(Effect=="fixed","Fixed",Effect)) %>% 
  mutate(Group=if_else(Group=="block","Block",Group)) %>% 
  mutate(Term=gsub("patch_type","Patch type: ",Term)) %>% 
  mutate(Term=gsub("cor__","correlation: ",Term)) %>% 
  mutate(Term=gsub("sd__","SD ",Term)) %>% 
  mutate(Term=gsub("[.]"," x ",Term)) %>% 
  mutate(SE=round(SE,3)) %>% 
  mutate(Estimate=round(Estimate,3)) %>% 
  mutate(Statistic=round(Statistic,3)) %>% 
  mutate(P=round(P,3)) %>% 
  mutate_all(as.character) %>% 
  mutate(P=if_else(as.numeric(P)<0.001,"< 0.001",P)) %>% 
  replace_na(list(SE="",Statistic="",P="")) %>%

  # need to round numbers above to 4, convert to character, then na_replace
  # replace_na(list(group=" ", std.error="",statistic=" ",p.value = "")) %>% 
# M0<-read_csv("~/Rstudio/corridor_beetles/corridor_docs/eric_ms_thesis/tables/M0.csv") %>% 
  kable(
      digits = 3,
      align="lcc",
      format="latex",
      row.names = FALSE,
      booktabs=T,
      linesep = "", #removes the blank line after every 5 lines
      caption = "Model of species abundance by patch type.") %>%
  kable_styling(bootstrap_options = c("striped"),
                full_width = F,
                latex_options=c("scale_down","hold_position"),
                font_size = 12,
                position = "center") %>%
  # column_spec(1, italic = T) %>%
  row_spec(0, bold = T)
  # row_spec(1:15, hline_after = T)
M0
```
\elandscape

```{r Table-aovM51, echo=FALSE,message = FALSE,warning=FALSE}
lmerM51<-read_csv(here("corridor_docs","eric_ms_thesis","tables","lmerM5_1.csv")) %>%
    mutate(term=case_when(
    term == "patchConnected" ~ "Patch: Connected",
    term == "patchRectangle" ~ "Patch: Rectangle",
    term == "patchWinged" ~ "Patch: Winged",
    term == "sd__(Intercept)" ~ "SD (Intercept)",
    term == "sd__Observation" ~ "SD (Observation)",
    .default = as.character(term))) |>
mutate(effect=if_else(effect=="ran_pars","Random",effect)) %>% 
  mutate(effect=if_else(effect=="fixed","Fixed",effect)) %>% 
  mutate(group=if_else(effect=="block","Block",group)) %>% 
  mutate(std.error=round(std.error,3)) %>% 
  mutate(statistic=round(statistic,3)) %>% 
  mutate(std.error=(as.character(std.error))) %>% 
  mutate(statistic=(as.character(statistic))) %>% 
  replace_na(list(group="",std.error="",statistic="")) %>% 
  rename(SE = std.error,
         Estimate=estimate,
         Term=term,
         Group=group,
         Effect=effect,
         Statistic=statistic) %>% 
  kable(
      digits = 3,
      align="lllccc",
      format="latex",
      row.names = FALSE,
      booktabs=T,
      linesep = "", #removes the blank line after every 5 lines
      caption = "Linear mixed effects model of pooled total biomass by patch type including sampling block as a random effect.") %>%
  kable_styling(bootstrap_options = c("striped"),
                full_width = F,
                latex_options=c("scale_down","hold_position"),
                font_size = 12,
                position = "center") %>%
  # column_spec(1, italic = T) %>%
  row_spec(0, bold = T)
  # row_spec(1:15, hline_after = T)
lmerM51
```
\newpage

@@

\blandscape
```{r Table-M3, echo=FALSE,message = FALSE,warning=FALSE}

@@

  # need to round numbers above to 4, convert to character, then na_replace
  # replace_na(list(group=" ", std.error="",statistic=" ",p.value = "")) %>% 
# M3<-read_csv("~/Rstudio/corridor_beetles/corridor_docs/eric_ms_thesis/tables/M3.csv") %>% 
  kable(
      digits = 3,
      align="lcc",
      format="latex",
      row.names = FALSE,
      booktabs=T,
      linesep = "", #removes the blank line after every 5 lines
      caption = "Generalized linear mixed model of abundance by patch type of the 6 most abundant species and interaction term for patch type and species including sampling block as a random effect.") %>%
  kable_styling(bootstrap_options = c("striped"),
                full_width = F,
                latex_options=c("scale_down","hold_position"),
                font_size = 12,
                position = "center") %>%
  # column_spec(1, italic = T) %>%
  row_spec(0, bold = T)
  # row_spec(1:15, hline_after = T)
M3
```
\elandscape
\newpage

@@

\vspace*{1cm}

\begin{center}
\textbf{BIOGRAPHICAL SKETCH}
\end{center}

<!-- _A biographical sketch is required of all candidates. The biographical sketch should be in narrative form. Third person, past tense, it typically includes the educational background of the candidate. The author should have replaced this paragraph with their own._ -->

\doublespacing

Eric Escobar-Chena completed his Bachelors education at Virginia Commonwealth University in 2023. During his time there he developed a fondness for insects which grew into a curiosity of the natural world. He later began to explore this curiosity deeper in beginning his graduate education at the University of Florida as a Master's Student under the supervision of Emilio Bruna. He received his Master of Science from the University of Florida in 2025.

>>>>

---
title             : "Effect of experimental habitat connectivity on dung beetle communities"
shorttitle        : "dung beetles and habitat corridors"

author: 
  - name          : "Eric Escobar-Chena"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "University of Florida"
    email         : "eescobarchena@ufl.edu"
    role: # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - Conceptualization
      - Methodology
      - Investigation
      - Formal analysis
      - Data curation
      - Software
      - Visualization
      - Writing - Original Draft Preparation
      - Writing - Review & Editing
  - name          : "Julian Resasco"
    affiliation   : "2"

    role:
      - Conceptualization
      - Methodology
      - Writing - Original Draft Preparation
      - Writing - Review & Editing

  - name          : "Emilio M. Bruna"
    affiliation   : "1,3"
    role:
      - Conceptualization
      - Methodology
      - Investigation
      - Formal analysis
      - Data curation
      - Visualization
      - Writing - Original Draft Preparation
      - Writing - Review & Editing
      - Project administration
      - Supervision

affiliation:
  - id            : "1"
    institution   : "Department of Wildlife Ecology & Conservation, University of Florida, PO Box 110530, Gainesville, Florida, 32611-0430 USA"
  - id            : "2"
    institution   : "University of Colorado, Boulder"
  - id            : "3"
    institution   : "Center for Latin American Studies, University of Florida, PO Box 110530, Gainesville, Florida, 32611-0530 USA"

note:       | 
  A preprint of this article has been posted at ----- Preprints (link).

  ORCID:
    EEC: ---- |
    EB: 0000-0003-3381-8477 |
    JR: ----- |

authornote: |
  <!-- EEC: Department of Wildlife Ecology & Conservation, University of Florida, PO Box 110530, Gainesville, Florida, 32611-0430 USA -->
  <!-- JR: University of Colorado, Boulder -->
  <!-- EB: Department of Wildlife Ecology & Conservation, University of Florida, PO Box 110530, Gainesville, Florida, 32611-0430 USA and Center for Latin American Studies, University of Florida, PO Box 110530, Gainesville, Florida, 32611-0530 USA -->

    We thank the USDA Forest Service for maintaining experimental landscapes and assisting in getting established at the site. We also wanted to specifically thank Thomas Smith for his help in data collection, Sara Escobar-Chena for her help in processing and data entry. Funding was provided by -----.

abstract: |

  Habitat fragmentation threatens biodiversity across the globe as habitat loss, isolation, and edge effects become increasingly prevalent. Corridors have become an important tool in order to combat the negative effects of fragmentation, however they are difficult to study in natural systems without incurring confounding effects. To observe changes in insect community composition as an effect of landscape features we sampled dung beetles in a landscape scale experiment. We did not see a difference in species richness or diversity, but dung beetle abundances were higher in continuous forest habitat and open habitat patches connected by a corridors than in isolated patches.

  <!-- https://tinyurl.com/ybremelq -->

keywords          : "keywords"
wordcount         : "4437"

bibliography      : "eric-thesis.bib"
csl               : "ecology.csl"
floatsintext      : no
linenumbers       : yes
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no
replace_ampersands: yes

classoption       : "man, donotrepeattitle"
output            : papaja::apa6_pdf

fig_caption: yes #yes will automatically add "Figure X" to the start 
# of the caption and reorders automatically if the figure position changes
always_allow_html: yes

header-includes:
  - \raggedbottom
  - \usepackage{endfloat} #[nomarkers] excludes the {insert figure x around here] from main text. The others exclude the list of tables and figures. https://cs.brown.edu/about/system/managed/latex/doc/endfloat.pdf
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{tabu}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
  - \usepackage{sectsty}\allsectionsfont{\raggedright}

---

```{r setup, include = FALSE}
library("papaja")

# r_refs("
# here("corridor_docs","eric_ms_thesis","eric-thesis.bib")
# ")

@@

@@

## INTRODUCTION

@@

## METHODS

### Study site

@@

## RESULTS

@@

## DISCUSSION

<!-- Paragraph 1: Overview: While there may indeed be effects of patch area and connectivity on dung beetle abundance on community composition, other (landscape-scale) drivers appear to be more important.  -->

@@

@@

Moving forward, there is a bounty of knowledge yet to be collected for dung beetles in the south eastern United States. Relative to what is already known in tropical ecosystems, this study provides a glimpse on how dung beetles are responding to fragmentation and connectivity in subtropic pine habitats, but much can still be learned about dung beetles in these spaces and the role they play in ecosystems as well as how they interact with landscapes. The unique role of dung beetles as waste removers is fascinating enough and of great importance for maintaining landscapes so dominated by pastures like Southern US [@cheng_dweller_2022; @WOS:000368243300001]. Not only waste removal but also secondary effect such as seed dispersal and parasitic reduction are important to understand in the context of fragmentation [@fincher_effects_1975; @vulinec_dung_2002]. SRS provides an excellent experimental design for direct comparison of landscapes but another main goal of the site is to study dispersal, dung beetles would be an excellent system for studying movement. They already show promise, in preliminary trials I released beetles from both the connected and rectangular patches with baited traps in the central patch. I recovered one recapture which originated from the connected patch, indicating that individuals do move through corridors.

@@

## LTERATURE CITED

@@

@@

\newpage
\blandscape

```{r Table-MO, echo=FALSE,message = FALSE,warning=FALSE}
M0<-read_csv(here("corridor_docs","eric_ms_thesis","tables","M0.csv")) %>%
  replace_na(list(group=" ")) %>% 
  rename(SE = std.error,
         Estimate=estimate,
         Term=term,
         Group=group,
         Effect=effect,
         Statistic=statistic,
         P=p.value) %>% 
  mutate(Effect=if_else(Effect=="ran_pars","Random",Effect)) %>% 
  mutate(Effect=if_else(Effect=="fixed","Fixed",Effect)) %>% 
  mutate(Group=if_else(Group=="block","Block",Group)) %>% 
  mutate(Term=gsub("patch_type","Patch type: ",Term)) %>% 
  mutate(Term=gsub("cor__","correlation: ",Term)) %>% 
  mutate(Term=gsub("sd__","SD ",Term)) %>% 
  mutate(Term=gsub("[.]"," x ",Term)) %>% 
  mutate(SE=round(SE,3)) %>% 
  mutate(Estimate=round(Estimate,3)) %>% 
  mutate(Statistic=round(Statistic,3)) %>% 
  mutate(P=round(P,3)) %>% 
  mutate_all(as.character) %>% 
  mutate(P=if_else(as.numeric(P)<0.001,"< 0.001",P)) %>% 
  replace_na(list(SE="",Statistic="",P="")) %>%

  # need to round numbers above to 4, convert to character, then na_replace
  # replace_na(list(group=" ", std.error="",statistic=" ",p.value = "")) %>% 
# M0<-read_csv("~/Rstudio/corridor_beetles/corridor_docs/eric_ms_thesis/tables/M0.csv") %>% 
  kable(
      digits = 3,
      align="lcc",
      format="latex",
      row.names = FALSE,
      booktabs=T,
      linesep = "", #removes the blank line after every 5 lines
      caption = "Model of species abundance by patch type.") %>%
  kable_styling(bootstrap_options = c("striped"),
                full_width = F,
                latex_options=c("scale_down","hold_position"),
                font_size = 12,
                position = "center") %>%
  # column_spec(1, italic = T) %>%
  row_spec(0, bold = T)
  # row_spec(1:15, hline_after = T)
M0
```

\elandscape
\newpage

```{r Table-aovM51, echo=FALSE,message = FALSE,warning=FALSE}
lmerM51<-read_csv(here("corridor_docs","eric_ms_thesis","tables","lmerM5_1.csv")) %>%
    mutate(term=case_when(
    term == "patchConnected" ~ "Patch: Connected",
    term == "patchRectangle" ~ "Patch: Rectangle",
    term == "patchWinged" ~ "Patch: Winged",
    term == "sd__(Intercept)" ~ "SD (Intercept)",
    term == "sd__Observation" ~ "SD (Observation)",
    .default = as.character(term))) |>
mutate(effect=if_else(effect=="ran_pars","Random",effect)) %>% 
  mutate(effect=if_else(effect=="fixed","Fixed",effect)) %>% 
  mutate(group=if_else(effect=="block","Block",group)) %>% 
  mutate(std.error=round(std.error,3)) %>% 
  mutate(statistic=round(statistic,3)) %>% 
  mutate(std.error=(as.character(std.error))) %>% 
  mutate(statistic=(as.character(statistic))) %>% 
  replace_na(list(group="",std.error="",statistic="")) %>% 
  rename(SE = std.error,
         Estimate=estimate,
         Term=term,
         Group=group,
         Effect=effect,
         Statistic=statistic) %>% 
  kable(
      digits = 3,
      align="lllccc",
      format="latex",
      row.names = FALSE,
      booktabs=T,
      linesep = "", #removes the blank line after every 5 lines
      caption = "Linear mixed effects model of pooled total biomass by patch type including sampling block as a random effect.") %>%
  kable_styling(bootstrap_options = c("striped"),
                full_width = F,
                latex_options=c("scale_down","hold_position"),
                font_size = 12,
                position = "center") %>%
  # column_spec(1, italic = T) %>%
  row_spec(0, bold = T)
  # row_spec(1:15, hline_after = T)
lmerM51
```

@@

@@

\newpage
\blandscape

```{r Table-M3, echo=FALSE,message = FALSE,warning=FALSE}

@@

  # need to round numbers above to 4, convert to character, then na_replace
  # replace_na(list(group=" ", std.error="",statistic=" ",p.value = "")) %>% 
# M3<-read_csv("~/Rstudio/corridor_beetles/corridor_docs/eric_ms_thesis/tables/M3.csv") %>% 
  kable(
      digits = 3,
      align="lcc",
      format="latex",
      row.names = FALSE,
      booktabs=T,
      linesep = "", #removes the blank line after every 5 lines
      caption = "Generalized linear mixed model of abundance by patch type of the 6 most abundant species and interaction term for patch type and species including sampling block as a random effect.") %>%
  kable_styling(bootstrap_options = c("striped"),
                full_width = F,
                latex_options=c("scale_down","hold_position"),
                font_size = 12,
                position = "center") %>%
  # column_spec(1, italic = T) %>%
  row_spec(0, bold = T)
  # row_spec(1:15, hline_after = T)
M3
```

\elandscape
\newpage

@@

```{r, label = patch, echo = FALSE, message = FALSE, warning = FALSE, fig.cap = "Average total dung beetle abundance by patch type with standard deviations as error bars.", out.width="70%", fig.align='center'}

\begin{center}
\end{center}
